DRAFT: Harmonisation and trait aggregation paper

### 2020-12-17 Paper/Draft_harmonization_paper.tex
17.12: Minor Edits Ralf

Synthesizing trait information from multiple trait databases is crucial to unfold the full potential of trait-based approaches, e.g. for studying community trait responses to environmental gradients worldwide. Our attempt of harmonising invertebrate trait information from trait databases of different regions showed that grouping feature harmonisation is labour-intensive because grouping features are differentiated differently into traits between trait databases and various codings are used to describe traits. In addition, the same traits are sometimes defined differently in the investigated databases, requiring expert knowledge to reclassify these traits (e.g. the trait piercer in the Tachet database, Table \ref{tab:traits_harmonisation}). Although, others have noted the lack of standardised trait terminology in freshwater ecology \cite{baird_toward_2011, brink_traits-based_2011} we are the first to describe invertebrate trait definition discrepancies for some commonly used grouping features.
%In addition we provide harmonised grouping feature trait datasets that can be used for other studies
To resolve definition discrepancies and facilitate data synthesis in the future terminological standards are needed. Harmonised definitions and concepts of traits have been developed in the past for other organism groups, e.g. for plants with the \textit{Thesaurus Of Plant characteristics} (TOP) initiative \cite{garnier_towards_2017}. The core of this harmonisation initiative is to provide standardised trait definitions and to draw connections to synonyms, related terms, and surrounding concepts by linking to other controlled vocabularies or ontologies in the field. Such initiatives could be a role model for freshwater ecologists to establish unambiguous terminologies for invertebrate traits. The existing freshwater invertebrate trait databases could be linked through standardised terminologies or ontologies, as suggested by Baird et al. \cite{baird_toward_2011} almost a decade ago. By following the recently proposed \textit{Ecological Trait-data Standard Vocabulary} (ETS) providers of invertebrate trait data could connect their traits to such ontologies \cite{schneider_towards_2019}. Once standardised trait definitions are established, these will improve trait data sharing, trait data processing when working with multiple trait databases, and ultimately interpretation of the derived results. % What is not mentioned here is how traits are measured, for which measurement protocols exist I believe. I think this is a separate issue.

@@

When trait values are only available at species-level and observational data are on less precise taxonomic levels various trait aggregation methods have been used by researchers. We aggregated traits from an Australian and North American trait dataset to family-level using 5 different aggregation methods and compared the results to traits assigned at family-level for these regions. Evaluation of the differences between aggregated and assigned traits is difficult because it remains unclear what the true value of a particular trait for a particular family is. Aggregation of trait information on species or genus-level to a point estimate on family-level suggests a precision which is not necessarily present, especially not for traits with a high variability or if trait information on species-level is missing. Some traits can vary strongly on more precise taxonomic levels than the family-level. For example, Monaghan and Soares \cite{monaghan_improving_2013} found a high intra-family trait dispersion in the Tachet database for traits of the grouping features body size, flow preference, and reproduction. Furthermore, studies focusing on the lability of traits, i.e. how much traits are unconstrained by phylogeny, found that traits of ecological preferences (e.g. thermal preference), body size, resistance forms, and to a lesser extent feeding mode are labile, and thus possibly highly variable \cite{poff_functional_2006, wilkes_traitbased_2020}. However, if trait aggregation is necessary our study can give guidance, under which circumstances the choice of aggregation method might be important. The 5 aggregation methods we tested use either the mean or the median and different weightings. Our results indicated that 1) the median aggregation methods compared to mean aggregation methods were often closer to the assigned traits at family-level and 2) that the different weightings tested exerted a minor impact on the aggregated trait affinities. We obtained for both datasets more differing cases for mean aggregation methods, irrespective of the weighting. In the Australian dataset, the amount of differing cases was slightly higher for the mean aggregation methods (23 \%) than for median aggregation methods (16 to 18 \% differing cases). In the North American dataset, mean aggregation methods yielded in 47 \% of the cases a difference to the assigned traits, much higher in comparison to the median aggregation methods (between 15 and 18 \%, Table \ref{tab:summary_stat_aggr_vs_fam_assigned}). When differences occurred between median aggregation methods and assigned traits, these differences were greater compared to the mean aggregation methods, particularly for the North American dataset. There, the mean absolute differences for the median aggregations were twice as high (0.63 - 0.7) compared to the mean aggregations (0.3 - 0.31, Table \ref{tab:summary_stat_aggr_vs_fam_assigned}). The differences can be explained by the binary coding used in the North American trait dataset and in the assigned traits. As a result, traits have affinities of either 1 or 0. Therefore, traits of a particular grouping feature have in most cases a higher difference in trait affinities to each other than if they were fuzzy coded. Using the median to aggregate binary coded taxa will often result in a value of 0 or 1, while using the mean will often result in values between 0 and 1. Thus, using the median led for the North American dataset either to agreement with the assigned traits or to higher differences compared to the mean aggregation methods.

The weightings explored in this analysis should lead to different aggregated trait affinities in cases where genera within a family have different numbers of species. In fact, the simulation results indicated that for a particular family where one genus has a much larger number of species compared to the other genera, weighting the number of taxa can lead to different aggregated trait affinities, especially for traits with higher levels of variance (Figure \ref{fig:sim_indv_runs}. However, in the comparison of the trait aggregation methods to assigned traits the number of differing cases and the mean absolute differences in trait affinities were similar across the mean aggregation and across the median aggregation methods, which suggests a small influence of the weightings on the aggregation methods. Also, the distributions of absolute trait affinity differences to assigned traits were relatively similar for the mean aggregation methods and the median aggregation methods respectively (Figure \ref{fig:diff_aggr_traits_chessman} and Figure \ref{fig:diff_aggr_traits_pyne}). The minor impact on trait aggregation of the weightings may be explained by the fact that a considerable portion of taxa had low numbers of genera or species entries. From the taxa that were compared from the North American trait dataset, 14 \% were resolved on family-level and 62 \% were resolved on genus-level. 52 \% had 5 or less genera and 13 \% contained just one genus (Figure \ref{fig:tax_hierarchy_NOA}). In the Australian dataset, 21 \% of the taxa were resolved on family-level, 40 \% were resolved on genus level, and 39 \% were resolved on species-level. 68 \% of the taxa contain 5 or less genera and 40 \% just contained one genus (Figure \ref{fig:tax_hierarchy_AUS}). Hence, these results might change when more trait information on species-level becomes available.

@@

We explored how using harmonised grouping features and aggregated traits might change the results of trait-environment relationships. We are not aware of other studies that compared trait-environment relationships using harmonised and not harmonised data. The re-analysis yielded only slightly different results compared to the original study, i.e. some of the traits that responded in the original analysis did not respond in the re-analysis. Those traits that were non-responsive in the re-analysis were those closest to the criterion that defined when a trait was associated with either high or low salinity (mahalanobis distance greater than the 97.5 \% quantile of the Chi-square distribution, see method section), namely feeding mode shredder, respiration gills, and life cycle duration traits. Feeding mode shredder was the only trait that was non-responsive when using harmonised (but not aggregated) as well as when using aggregated data. One reason for the non-responsiveness of shredders in the re-analysis was likely the harmonisation procedure. During harmonisation the trait shredder has been amalgamated based on three traits (Miner, xylophagus, and shredder). As a result, the trait affinities in the original data have a higher mean and standard deviation than in the aggregated and harmonised data (Table \ref{tab:SI_resp_traits_summary_stats}) suggesting that the signal in the original data has presumably been weakened by the harmonisation. This might be the reason that shredders showed a less strong reaction to the salinity gradient in the re-analysis. These findings show that if harmonisation is necessary, harmonised and non-harmonised data, if available, should be compared and possible averaging effects should be considered in further analysis. The fact that the family-level aggregated traits used in the re-analysis showed only slightly different results compared to the original analysis is in line with the findings of others that family-level traits can be sensitive enough to detect environmental impacts \cite{beketov_spear_2009}. % Cite another paper that uses family-level trait data
% Although species-level traits are expected to result in more precise results when investigating trait-environment relationships. % Cite Schmidt Kloiber?

@@

With the increasing use of computer vision in ecology, species identification could become easier and more automated, and trait aggregation might be less needed \cite{milosevic_application_2020}. However, if traits need to be aggregated we showed that trait aggregation is often in conclusion with expert assignments, specifically aggregation approaches using the median. Which aggregation method the objectively better measure is cannot be answered, but when trait aggregation is necessary 1) highly variable traits and taxa where a few genera have a high number of species should be treated carefully, 2) coding of traits should be considered, and 3) weighting trait information based on the taxa present in the used trait database does not seem to have much influence. No matter which aggregation approach is chosen, for future studies in which researchers must aggregate traits, a measure of trait dispersion should be reported to indicate the uncertainty of the aggregated estimate.

>>>>

Synthesizing trait information from multiple trait databases is crucial to unfold the full potential of trait-based approaches, e.g. for studying community trait responses to environmental gradients worldwide. Our attempt of harmonising invertebrate trait information from trait databases of different regions showed that grouping feature harmonisation is labour-intensive because grouping features are differentiated differently into traits between trait databases and various codings are used to describe traits. In addition, the same traits are sometimes defined differently in the investigated databases, requiring expert knowledge to reclassify these traits (e.g. the trait piercer in the Tachet database, Table \ref{tab:traits_harmonisation}). Although, others have noted the lack of standardised trait terminology in freshwater ecology \cite{baird_toward_2011, brink_traits-based_2011}, we are the first to describe invertebrate trait definition discrepancies for some commonly used grouping features.
%In addition we provide harmonised grouping feature trait datasets that can be used for other studies
To resolve definition discrepancies and facilitate data synthesis in the future,  terminological standards are needed. Harmonised definitions and concepts of traits have been developed in the past for other organism groups, e.g. for plants with the \textit{Thesaurus Of Plant characteristics} (TOP) initiative \cite{garnier_towards_2017}. The core of this harmonisation initiative is to provide standardised trait definitions and to draw connections to synonyms, related terms, and surrounding concepts by linking to other controlled vocabularies or ontologies in the field. Such initiatives could be a role model for freshwater ecologists to establish unambiguous terminologies for invertebrate traits. The existing freshwater invertebrate trait databases could be linked through standardised terminologies or ontologies, as suggested by Baird et al. \cite{baird_toward_2011} almost a decade ago. By following the recently proposed \textit{Ecological Trait-data Standard Vocabulary} (ETS) providers of invertebrate trait data could connect their traits to such ontologies \cite{schneider_towards_2019}. Once standardised trait definitions are established, these will improve trait data sharing, trait data processing when working with multiple trait databases, and ultimately interpretation of the derived results. % What is not mentioned here is how traits are measured, for which measurement protocols exist I believe. I think this is a separate issue.

@@

When trait values are only available at species-level and observational data are on less precise taxonomic levels various trait aggregation methods have been used by researchers. We aggregated traits from an Australian and North American trait dataset to family-level using 5 different aggregation methods and compared the results to traits assigned at family-level for these regions. Evaluation of the differences between aggregated and assigned traits is difficult because it remains unclear what the true value of a particular trait for a particular family is. Aggregation of trait information on species or genus-level to a point estimate on family-level suggests a precision that is not necessarily present, especially not for traits with a high variability or if trait information on species-level is missing. Some traits can vary strongly on more precise taxonomic levels than the family-level. For example, Monaghan and Soares \cite{monaghan_improving_2013} found a high intra-family trait dispersion in the Tachet database for traits of the grouping features body size, flow preference, and reproduction. Furthermore, studies focusing on the lability of traits, i.e. how much traits are unconstrained by phylogeny, found that traits of ecological preferences (e.g. thermal preference), body size, resistance forms, and to a lesser extent feeding mode are labile, and thus possibly highly variable \cite{poff_functional_2006, wilkes_traitbased_2020}. However, if trait aggregation is necessary our study can give guidance, under which circumstances the choice of aggregation method might be important. The 5 aggregation methods we tested use either the mean or the median and different weightings. Our results indicated that 1) the median aggregation methods compared to mean aggregation methods were often closer to the assigned traits at family-level and 2) that the different weightings tested exerted a minor impact on the aggregated trait affinities. We obtained for both datasets more differing cases for mean aggregation methods, irrespective of the weighting. In the Australian dataset, the amount of differing cases was slightly higher for the mean aggregation methods (23 \%) than for median aggregation methods (16 to 18 \% differing cases). In the North American dataset, mean aggregation methods yielded in 47 \% of the cases a difference to the assigned traits, much higher in comparison to the median aggregation methods (between 15 and 18 \%, Table \ref{tab:summary_stat_aggr_vs_fam_assigned}). When differences occurred between median aggregation methods and assigned traits, these differences were greater compared to the mean aggregation methods, particularly for the North American dataset. There, the mean absolute differences for the median aggregations were twice as high (0.63 - 0.7) compared to the mean aggregations (0.3 - 0.31, Table \ref{tab:summary_stat_aggr_vs_fam_assigned}). The differences can be explained by the binary coding used in the North American trait dataset and in the assigned traits. As a result, traits have affinities of either 1 or 0. Therefore, traits of a particular grouping feature have in most cases a higher difference in trait affinities to each other than if they were fuzzy coded. Using the median to aggregate binary coded taxa will often result in a value of 0 or 1, while using the mean will often result in values between 0 and 1. Thus, using the median led for the North American dataset either to agreement with the assigned traits or to higher differences compared to the mean aggregation methods.

The weightings explored in this analysis should lead to different aggregated trait affinities in cases where genera within a family have different numbers of species. In fact, the simulation results indicated that for a particular family where one genus has a much larger number of species compared to the other genera, weighting the number of taxa can lead to different aggregated trait affinities, especially for traits with higher levels of variance (Figure \ref{fig:sim_indv_runs}). However, in the comparison of the trait aggregation methods to assigned traits the number of differing cases and the mean absolute differences in trait affinities were similar across the mean aggregation and across the median aggregation methods, which suggests a small influence of the weightings on the aggregation methods. Also, the distributions of absolute trait affinity differences to assigned traits were relatively similar for the mean aggregation methods and the median aggregation methods, respectively (Figure \ref{fig:diff_aggr_traits_chessman} and Figure \ref{fig:diff_aggr_traits_pyne}). The minor impact on trait aggregation of the weightings may be explained by the fact that a considerable portion of taxa had low numbers of genera or species entries. From the taxa that were compared from the North American trait dataset, 14 \% were resolved on family-level and 62 \% were resolved on genus-level. 52 \% had 5 or less genera and 13 \% contained just one genus (Figure \ref{fig:tax_hierarchy_NOA}). In the Australian dataset, 21 \% of the taxa were resolved on family-level, 40 \% were resolved on genus level, and 39 \% were resolved on species-level. 68 \% of the taxa contain 5 or less genera and 40 \% just contained one genus (Figure \ref{fig:tax_hierarchy_AUS}). Hence, these results might change when more trait information on species-level becomes available.

@@

We explored how using harmonised grouping features and aggregated traits might change the results of trait-environment relationships. We are not aware of other studies that compared trait-environment relationships using harmonised and non harmonised data. The re-analysis yielded only slightly different results compared to the original study, i.e. some of the traits that responded in the original analysis did not respond in the re-analysis. Those traits that were non-responsive in the re-analysis were those closest to the criterion that defined when a trait was associated with either high or low salinity (Mahalanobis distance greater than the 97.5 \% quantile of the Chi-square distribution, see method section), namely feeding mode shredder, respiration gills, and life cycle duration traits. Feeding mode shredder was the only trait that was non-responsive when using harmonised (but not aggregated) as well as when using aggregated data. One reason for the non-responsiveness of shredders in the re-analysis was likely the harmonisation procedure. During harmonisation the trait shredder has been amalgamated based on three traits (miner, xylophagus, and shredder). As a result, the trait affinities in the original data have a higher mean and standard deviation than in the aggregated and harmonised data (Table \ref{tab:SI_resp_traits_summary_stats}) suggesting that the signal in the original data has presumably been weakened by the harmonisation. This might be the reason that shredders showed a less strong reaction to the salinity gradient in the re-analysis. These findings show that if harmonisation is necessary, harmonised and non-harmonised data, if available, should be compared and possible averaging effects should be considered in further analysis. The fact that the family-level aggregated traits used in the re-analysis showed only slightly different results compared to the original analysis is in line with the findings of others that family-level traits can be sensitive enough to detect environmental impacts \cite{beketov_spear_2009}. % Cite another paper that uses family-level trait data
% Although species-level traits are expected to result in more precise results when investigating trait-environment relationships. % Cite Schmidt Kloiber?

@@

With the increasing use of computer vision in ecology, both species identification and trait quantification could be automated \cite{milosevic_application_2020, hoye_deep_2020}. This might result in more observational data and trait data on species-level in the future. However, if traits need to be aggregated we showed that trait aggregation is often in conclusion with expert assignments, specifically aggregation approaches using the median. Which aggregation method the objectively better measure is cannot be answered, but when trait aggregation is necessary 1) median aggregation seems to be favorable, 2) highly variable traits and taxa where a few genera have a high number of species should be treated carefully, 3) coding of traits should be considered, and 4) weighting trait information based on the taxa present in the used trait database does not seem to have much influence. No matter which aggregation approach is chosen, for future studies in which researchers must aggregate traits, a measure of trait dispersion should be reported to indicate the uncertainty of the aggregated estimate.
